# S2N R package

# Introduction
Existing methods have been proposed to identify sample-specific network (a network for a sample) from bulk and single-cell transcriptomics data with different programming languages (e.g. Matlab, Python, R). To facilitate comparison and the selection of a more suitable method for exploring sample-specific network in transcriptomics data, we implement S2N R package.

# Install S2N R package

```{r, eval=TRUE, include=TRUE}
install.packages("devtools")
library(devtools)
install_github("zhangjunpeng411/S2N")
```
# Load BRCA sample data
The BRCA sample data contains the following:
1. Bulk RNA-seq data, single-cell RNA-seq (scRNA-seq) data, and spatial transcriptomic (ST) data. 
2. Mutation data.
3. Putative ncRNA regulatory network.
4. Hub ncRNAs.

```{r, eval=TRUE, include=TRUE}
data(bulk_ncR_tar_normal_cancer)
data(scRNA_ncR_tar_normal_cancer)
data(scRNA_ncR_tar_count_cancer)
data(ST_ncR_tar_cancer)
data(ST_ncR_tar_count_cancer)
data(mut_ncR)
data(network)
data(hub_ncRNAs)
```

# Inferring sample-specific network
S2N provides six computational methods to identify sample-specific network, including SSN, Paired-SSN, LIONESS, SCS, CSN, and Scan.perturb. We provide a quick example with SSN to run S2N.

```{r echo=FALSE, results='hide', message=FALSE}
library(S2N)

# bulk data
bulk_SSN_res <- SSN(bulk_ncR_normal[1:200, ], bulk_tar_normal[1:500, ], bulk_ncR_cancer[1:200, ], bulk_tar_cancer[1:500, ], cormethod = "pearson")

# scRNA data
scRNA_SSN_res <- SSN(scRNA_ncR_normal[1:100, 1:5], scRNA_tar_normal[1:1000, 1:5], scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:1000, 1:5], cormethod = "pearson")
```

# License
GPL-3

