---
title: "S2N: an R package to infer sample-specific network"
author: "Junpeng Zhang"
date: '`r Sys.Date()`'
output: 
    BiocStyle::html_document:
      toc: yes
    BiocStyle::pdf_document:
      toc: yes
vignette: >
  %\VignettePackage{S2N}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{S2N: an R package to infer sample-specific network}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
    warning = FALSE,
    message = FALSE)
```

```{r echo=FALSE, results='hide', message=FALSE}
library(S2N)
```

# Introduction
Existing methods have been proposed to identify sample-specific network (a network for a sample) from bulk and single-cell transcriptomics data with different programming languages (e.g. Matlab, Python, R). To facilitate comparison and the selection of a more suitable method for exploring sample-specific network in transcriptomics data, we implement S2N R package.

# Inferring sample-specific network
S2N provides six computational methods to identify sample-specific network, including Single-Sample Network (SSN) [1], Paired-Single-Sample Network (Paired-SSN) [2], Linear Interpolation to Obtain Network Estimates for Single Samples (LIONESS) [3], SCS (single-sample controller strategy) [4], CSN (cell-specific network) [5, 6], Scan.perturb (Sample-speCific miRNA regulAtioN with statistical perturbation strategy) [7]. 

## Loading BRCA sample data
The BRCA sample data contains the following:
1. Bulk RNA-seq data, single-cell RNA-seq (scRNA-seq) data, and spatial transcriptomic (ST) data. 
2. Mutation data.
3. Putative ncRNA regulatory network.
4. Hub ncRNAs.

```{r, eval=TRUE, include=TRUE}
data(bulk_ncR_tar_normal_cancer)
data(scRNA_ncR_tar_normal_cancer)
data(scRNA_ncR_tar_count_cancer)
data(ST_ncR_tar_cancer)
data(ST_ncR_tar_count_cancer)
data(mut_ncR)
data(network)
data(hub_ncRNAs)
```

## SSN
By using SSN method, we identify sample-specific ncRNA regulatory network from bulk RNA-seq or scRNA-seq data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_SSN_res <- SSN(bulk_ncR_normal[1:100, ], bulk_tar_normal[1:500, ], bulk_ncR_cancer[1:100, ], bulk_tar_cancer[1:500, ], cormethod = "pearson")

# scRNA data
scRNA_SSN_res <- SSN(scRNA_ncR_normal[1:100, 1:5], scRNA_tar_normal[1:500, 1:5], scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], cormethod = "pearson")
```

## Paired-SSN
We identify sample-specific ncRNA regulatory network by using Paired-SSN method from bulk RNA-seq or scRNA-seq data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_PaiedSSN_res <- PairedSSN(bulk_ncR_normal[1:100, ], bulk_tar_normal[1:500, ], bulk_ncR_cancer[1:100, ], bulk_tar_cancer[1:500, ], cormethod = "pearson")

# scRNA data
scRNA_PaiedSSN_res <- PairedSSN(scRNA_ncR_normal[1:100, 1:5], scRNA_tar_normal[1:500, 1:5], scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], cormethod = "pearson")
```

## LIONESS
By using the LIONESS method, we sample-specific ncRNA regulatory network from bulk RNA-seq, scRNA-seq, and ST data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_LIONESS_res <- LIONESS(bulk_ncR_cancer[1:100, 1:5], bulk_tar_cancer[1:500, 1:5], cormethod = "pearson")

# scRNA data
scRNA_LIONESS_res <- LIONESS(scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], cormethod = "pearson")

# ST data
ST_LIONESS_res <- LIONESS(ST_ncR_cancer[1:100, 1:5], ST_tar_cancer[1:500, 1:5], cormethod = "pearson")
```

## SCS
We adapt the SCS method to identify sample-specific ncRNA regulatory network from bulk RNA-seq, scRNA-seq, and ST data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_scs_res <- SCS(mut_ncR, bulk_ncR_cancer[1:100, 1:5], bulk_tar_cancer[1:500, 1:5], cormethod = "pearson", number = 10)

# scRNA data
scRNA_scs_res <- SCS(mut_ncR, scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], cormethod = "pearson", number = 10)

# ST data
ST_scs_res <- SCS(mut_ncR, ST_ncR_cancer[1:100, 1:5], ST_tar_cancer[1:500, 1:5], cormethod = "pearson", number = 10)
```

## CSN
The CSN method is used to identify sample-specific ncRNA regulatory network from bulk RNA-seq, scRNA-seq, and ST data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_CSN_res <- CSN(bulk_ncR_cancer[1:100, 1:5], bulk_tar_cancer[1:500, 1:5], boxsize = 0.1, p.value.cutoff = 0.05)

# scRNA data
scRNA_CSN_res <- CSN(scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], boxsize = 0.1, p.value.cutoff = 0.05)

# ST data
ST_CSN_res <- CSN(ST_ncR_cancer[1:100, 1:5], ST_tar_cancer[1:500, 1:5], boxsize = 0.1, p.value.cutoff = 0.05)
```

## Scan.perturb
The Scan.perturb method is used to identify sample-specific ncRNA regulatory network from bulk RNA-seq, scRNA-seq, and ST data.

```{r, eval=TRUE, include=TRUE}
# bulk data
bulk_Scan.perturb_res <- Scan.perturb(bulk_ncR_cancer[1:100, 1:5], bulk_tar_cancer[1:500, 1:5], method = "Pearson", p.value.cutoff = 0.05)

# scRNA data
scRNA_Scan.perturb_res <- Scan.perturb(scRNA_ncR_cancer[1:100, 1:5], scRNA_tar_cancer[1:500, 1:5], method = "Pearson", p.value.cutoff = 0.05)

# ST data
ST_Scan.perturb_res <- Scan.perturb(ST_ncR_cancer[1:100, 1:5], ST_tar_cancer[1:500, 1:5], method = "Pearson", p.value.cutoff = 0.05)
```

# Downstream analysis

## Validating sample-specific network
The function 'Network_validation' is implemented to validate sample-specific network.

## Identifying sample-specific hub regulators
The function 'Network_hub' is implemented to Identifying sample-specific hub regulators with the largest 20% degree in default.

## Predicting sample-specific modules
To identify sample-specific modules from sample-specifc network, we implement 'Network_module' function.

## Calculating sample-sample similarity matrix between two lists of sample-specific networks
The function 'Sim.network' is used to calculate sample-sample similarity matrix between two list of sample-specific networks.

## Performing hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific networks
The function 'Network_sim_hc' is implemented to perform hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific networks.

## Calculating sample-sample similarity matrix between two lists of sample-specific hubs
The function 'Sim.hub' is used to calculate sample-sample similarity matrix between two list of sample-specific hubs.

## Performing hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific hubs
The function 'Hub_sim_hc' is implemented to perform hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific hubs.

## Calculating sample-sample similarity matrix between two lists of sample-specific modules
The function 'Sim.module' is used to calculate sample-sample similarity matrix between two list of sample-specific modules.

## Performing hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific modules
The function 'Module_sim_hc' is implemented to perform hierarchical clustering analysis of samples using the calculated sample-sample similarity matrix in terms of sample-specific modules.

# Conclusions
S2N provides several built-in methods to infer and analyze sample-specific network, and it could provide a useful tool to explore sample-specific network from bulk RNA-seq, scRNA-seq, and ST data.

# References
[1] Liu X, Wang Y, Ji H, Aihara K, Chen L. Personalized characterization of diseases using sample-specific networks. Nucleic Acids Res. 2016;44(22):e164.

[2] Guo WF, Zhang SW, Zeng T, Li Y, Gao J, Chen L. A novel network control model for identifying personalized driver genes in cancer. PLoS Comput Biol. 2019;15(11):e1007520.

[3] Kuijjer ML, Tung MG, Yuan G, Quackenbush J, Glass K. Estimating Sample-Specific Regulatory Networks. iScience. 2019;14:226-240.

[4] Guo WF, Zhang SW, Liu LL, Liu F, Shi QQ, Zhang L, Tang Y, Zeng T, Chen L. Discovering personalized driver mutation profiles of single samples in cancer by network control strategy. Bioinformatics. 2018;34(11):1893-1903.

[5] Zhang J, Liu L, Xu T, Zhang W, Zhao C, Li S, Li J, Rao N, Le TD. Exploring cell-specific miRNA regulation with single-cell miRNA-mRNA co-sequencing data. BMC Bioinformatics.2021;22(1):578.

[6] Dai H, Li L, Zeng T, Chen L. Cell-specific network constructed by single-cell RNA sequencing data. Nucleic Acids Res. 2019;47(11):e62.

[7] Zhang J, Liu L, Wei X, Zhao C, Luo Y, Li J, Le TD. Scanning sample-specific miRNA regulation from bulk and single-cell RNA-sequencing data. BMC Biol. 2024;22(1):218.

# Session information
```{r}
sessionInfo()
```

