---
title: "Tutorial for inferring sample-specific miRNA regulatory networks with S2N in the K562 dataset" 
author: "\\
    
    Junpeng Zhang (zjp@dali.edu.cn)\\
    
    School of Engineering, Dali University"
date: '`r Sys.Date()`'
output:
    BiocStyle::html_document:
      toc: yes
    BiocStyle::pdf_document:
      toc: yes
vignette: >
    %\VignetteIndexEntry{Tutorial for inferring sample-specific miRNA regulatory networks with S2N in the K562 dataset} 
    %\VignetteEngine{knitr::rmarkdown} 
    %\usepackage[utf8]{inputenc} 
    %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
    warning = FALSE,
    message = FALSE)
```

# Introduction
In this tutorial, we apply eight built-in methods (LIONESS, SCS, CSN, Scan.perturb, Scan.interp, SWEET, SINUM, and BONOBO) of S2N to identify sample-specific miRNA regulatory networks from the K562 dataset (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114071). It is noted that SSN and Paired-SSN are unsuitable for the K562 dataset due to the absence of gene expression data in normal.

# Installation of S2N
S2N is freely available open-source under GNU GPL-3.0 from: https://github.com/zhangjunpeng411/S2N/, should be installed from GitHub prior to use.

```{r, eval=TRUE, include=TRUE}
install.packages("remotes")
library(remotes)
remotes::install_github("zhangjunpeng411/S2N")
library(S2N)
```

# Data preparation
The single-cell miRNA-mRNA co-sequencing data (GSE114071_NW_scsmRNA_K562_norm_log2.gct.gz for miRNA expression data, and GSE114071_NW_half_cell_K562_RNAseq_processed_data.gct.gz for mRNA expression data) is from the Gene Expression Omnibus (GEO, https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE114071). Please paste all downloaded files into a single folder (set the folder as the directory of R environment).

```{r, eval=TRUE, include=TRUE}
miR_data <- read.table("GSE114071_NW_scsmRNA_K562_norm_log2.gct", header = TRUE, sep='\t')
mR_data <-  read.table("GSE114071_NW_half_cell_K562_RNAseq_processed_data.gct", header = TRUE, skip = 2, sep='\t')
miR_names <- miR_data[, 1]
mR_names <- mR_data[, 1]
miR_data <- miR_data[, -seq(2)]
mR_data <- mR_data[, -seq(2)]
miR_data[is.na(miR_data)] <- 0
mR_data[is.na(mR_data)] <- 0
miR_names <- make.unique(miR_names)
mR_names <- make.unique(mR_names)
rownames(miR_data) <- miR_names
rownames(mR_data) <- mR_names
```

To ensure the consistency of the downloaded miRNA and mRNA expression data, the downloaded miRNA expression data is firstly processed with transformation. Moreover, the unmatched K562 cells between miRNA and mRNA expression data are removed. Finally, the matched miRNA and mRNA expression data are further normalized.

```{r, eval=TRUE, include=TRUE}
miR_data <- 2^(miR_data)
miR_scRNA_matched <- miR_data[, seq(19)] 
mR_scRNA_matched <- mR_data[, -c(6, 21, 22)]
miR_scRNA_norm <- log2(miR_scRNA_matched + 1)
mR_scRNA_norm <- log2(mR_scRNA_matched + 1)
```

The K562 dataset is further preprocessed by removing duplicate and invariant genes, retaining only highly expressed and variable miRNAs and mRNAs.

```{r, eval=TRUE, include=TRUE}
## Function for computing the average expression values of duplicate genes
# Exp_scRNA: Gene expression values of miRNAs or mRNAs in single cells, rows are cells and 
# columns are genes
# temp: Single cell expression data without duplicate genes
Averg_Duplicate <- function(Exp_scRNA){      
      raw_gene_names <- gsub("\\.\\d+$", "", rownames(Exp_scRNA))
      uniqueNameList <- unique(raw_gene_names)
      noOfgenes <- length(uniqueNameList)
      temp <- matrix(0, nrow = noOfgenes, ncol = ncol(Exp_scRNA))
      rownames(temp) <- uniqueNameList
      colnames(temp) <- colnames(Exp_scRNA)
      for(r in 1:noOfgenes){
          GeneList <- which(raw_gene_names == rownames(temp)[r])
          for(c in 1:ncol(temp)) {
              temp[r, c] <- mean(as.numeric(Exp_scRNA[GeneList, c]))  
          }
      }
    return(temp)
}

## Compute the average expression values of duplicate genes
miR_scRNA_average <- Averg_Duplicate(miR_scRNA_norm)
mR_scRNA_average <- Averg_Duplicate(mR_scRNA_norm)

## Remove genes with constant expression values in all cells
miR_scRNA_sd <- unlist(lapply(seq(dim(miR_scRNA_average)[1]), function(i) sd(miR_scRNA_average[i, ])))
mR_scRNA_sd <- unlist(lapply(seq(dim(mR_scRNA_average)[1]), function(i) sd(mR_scRNA_average[i, ])))
miR_scRNA_filter <- miR_scRNA_average[which(miR_scRNA_sd > 0), ]
mR_scRNA_filter <- mR_scRNA_average[which(mR_scRNA_sd > 0), ]

## Reserve high expression and variable miRNAs and mRNAs
miR_me <- unlist(lapply(seq(nrow(miR_scRNA_filter)), function(i) mean(miR_scRNA_filter[i, ])))
mR_me <- unlist(lapply(seq(nrow(mR_scRNA_filter)), function(i) mean(mR_scRNA_filter[i, ])))
miR_scRNA <- miR_scRNA_filter[which(miR_me > median(miR_me)), ]
mR_scRNA <- mR_scRNA_filter[which(mR_me > median(mR_me)), ]
```

# Sample-specific miRNA regulatory network construction
We use default parameters of eight built-in methods (LIONESS, SCS, CSN, Scan.perturb, Scan.interp, SWEET, SINUM, and BONOBO) to infer cell-specific miRNA regulatory networks. In this context, each cell represents a sample.

```{r, eval=TRUE, include=TRUE}
install.packages("peakRAM")
library(peakRAM)

dir.create("K562_results", showWarnings = FALSE, recursive = TRUE)

exprs <- list(
  LIONESS = quote({
    res_LIONESS <- LIONESS(miR_scRNA, mR_scRNA, num.cores = 16)
    saveRDS(res_LIONESS, "K562_results/scRNA_LIONESS_res.rds")
  }),
  SCS = quote({
    res_SCS <- SCS(rownames(miR_scRNA), miR_scRNA, mR_scRNA, num.cores = 16)
    saveRDS(res_SCS, "K562_results/scRNA_SCS_res.rds")
  }),
  CSN = quote({
    res_CSN <- CSN(miR_scRNA, mR_scRNA, bootstrap = TRUE, bootstrap_betw_point = 5, bootstrap_num = 10, num.cores = 16)
    saveRDS(res_CSN, "K562_results/scRNA_CSN_res.rds")
  }),
  Scan.perturb = quote({
    res_Scan.perturb <- Scan.perturb(miR_scRNA, mR_scRNA, method = 'pearson', num.cores = 16)
    saveRDS(res_Scan.perturb, "K562_results/scRNA_Scan.perturb_res.rds")
  }),
  Scan.interp = quote({
    res_Scan.interp <- Scan.interp(miR_scRNA, mR_scRNA, method = 'pearson', num.cores = 16)
    saveRDS(res_Scan.interp, "K562_results/scRNA_Scan.interp_res.rds")
  }),
  SWEET = quote({
    res_SWEET <- SWEET(miR_scRNA, mR_scRNA, num.cores = 16)
    saveRDS(res_SWEET, "K562_results/scRNA_SWEET_res.rds")
  }),
  SINUM = quote({
    res_SINUM <- SINUM(miR_scRNA, mR_scRNA, num.cores = 16)
    saveRDS(res_SINUM, "K562_results/scRNA_SINUM_res.rds")
  }),
  BONOBO = quote({
    res_BONOBO <- BONOBO(miR_scRNA, mR_scRNA, num.cores = 16)
    saveRDS(res_BONOBO, "K562_results/scRNA_BONOBO_res.rds")
  })
)

peak_results <- do.call(peakRAM, exprs)
saveRDS(peak_results, "K562_results/peakRAM_full_metrics_scRNA.rds")
```

# Sample-specific miRNA regulatory network validation
To validate the predicted sample-specific miRNA regulatory networks, we use established ground truth from miRTarBase v10.0 and TarBase v9.0.

```{r, eval=TRUE, include=TRUE}
data(miRTarget_Groundtruth)
scRNA_LIONESS_validation <- Network_validation(res_LIONESS, miRTarget_Groundtruth, calculate_metrics = TRUE)
scRNA_CSN_validation <- Network_validation(res_CSN, miRTarget_Groundtruth, calculate_metrics = TRUE)
scRNA_Scan.perturb_validation <- Network_validation(res_Scan.perturb, miRTarget_Groundtruth, calculate_metrics = TRUE)
scRNA_Scan.interp_validation <- Network_validation(res_Scan.interp, miRTarget_Groundtruth, calculate_metrics = TRUE)
scRNA_SWEET_validation <- Network_validation(res_SWEET, miRTarget_Groundtruth, calculate_metrics = TRUE)
scRNA_SINUM_validation <- Network_validation(res_SINUM, miRTarget_Groundtruth, calculate_metrics = TRUE)

# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract performance values from each validation list
df_list <- list(
  LIONESS      = scRNA_LIONESS_validation[[2]][,4]*100,
  CSN          = scRNA_CSN_validation[[2]][,4]*100,
  Scan.perturb = scRNA_Scan.perturb_validation[[2]][,4]*100,
  Scan.interp  = scRNA_Scan.interp_validation[[2]][,4]*100,
  SWEET        = scRNA_SWEET_validation[[2]][,4]*100,
  SINUM        = scRNA_SINUM_validation[[2]][,4]*100
)

# Convert to long-format data frame (Method, Value)
df_long <- bind_rows(lapply(names(df_list), function(method) {
  data.frame(Method = method, Value = df_list[[method]])
}))

# Set Method as factor with the desired order
df_long$Method <- factor(df_long$Method, levels = names(df_list))

# Nature-inspired color palette 
science_colors <- c("#0072B5", "#E18727", "#20854E", "#BC3C29", 
                    "#7876B1", "#6A3D9A")

# Generate boxplot with jittered points
p <- ggplot(df_long, aes(x = Method, y = Value, fill = Method)) +
  geom_boxplot(width = 0.6, alpha = 0.7, outlier.size = 0.5, 
               color = "black", outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, size = 0.5, alpha = 0.3, color = "black") +
  scale_fill_manual(values = science_colors[1:length(unique(df_long$Method))]) +
  labs(x = "", y = "Percentage of validated miRNA targets (%)") +
  theme_classic() +
  theme(text = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        axis.line = element_line(linewidth = 0.5, color = "black"),
        panel.grid = element_blank(),
        legend.position = "none")

# Display and save
print(p)
ggsave("performance_boxplot.png", p, width = 10, height = 9, dpi = 300, units = "cm")
```

# Sample similarity analysis
We calculate pairwise sample similarity based on three key features derived from each sample-specific network: (i) network architecture, (ii) hub gene composition, and (iii) modular organization. Histograms with kernel density overlay were generated from the K562 half-cell similarity matrices derived from sample-specific networks, hubs, and modules produced by the LIONESS method.

```{r, eval=TRUE, include=TRUE}
# network-level
scRNA_LIONESS_net_sim <- Sim.network(res_LIONESS,  res_LIONESS)
scRNA_CSN_net_sim <- Sim.network(res_CSN, res_CSN)
scRNA_Scan.perturb_net_sim <- Sim.network(res_Scan.perturb, res_Scan.perturb)
scRNA_Scan.interp_net_sim <- Sim.network(res_Scan.interp, res_Scan.interp)
scRNA_SWEET_net_sim <- Sim.network(res_SWEET, res_SWEET)
scRNA_SINUM_net_sim <- Sim.network(res_SINUM, res_SINUM)

# hub-level
scRNA_LIONESS_hub <- Network_hub(res_LIONESS)
scRNA_CSN_hub <- Network_hub(res_CSN)
scRNA_Scan.perturb_hub <- Network_hub(res_Scan.perturb)
scRNA_Scan.interp_hub <- Network_hub(res_Scan.interp)
scRNA_SWEET_hub <- Network_hub(res_SWEET)
scRNA_SINUM_hub <- Network_hub(res_SINUM)

scRNA_LIONESS_hub_sim <- Sim.hub(scRNA_LIONESS_hub,  scRNA_LIONESS_hub)
scRNA_CSN_hub_sim <- Sim.hub(scRNA_CSN_hub, scRNA_CSN_hub)
scRNA_Scan.perturb_hub_sim <- Sim.hub(scRNA_Scan.perturb_hub, scRNA_Scan.perturb_hub)
scRNA_Scan.interp_hub_sim <- Sim.hub(scRNA_Scan.interp_hub, scRNA_Scan.interp_hub)
scRNA_SWEET_hub_sim <- Sim.hub(scRNA_SWEET_hub, scRNA_SWEET_hub)
scRNA_SINUM_hub_sim <- Sim.hub(scRNA_SINUM_hub, scRNA_SINUM_hub)

# module-level
scRNA_LIONESS_module <- Network_module(res_LIONESS)
scRNA_CSN_module <- Network_module(res_CSN)
scRNA_Scan.perturb_module <- Network_module(res_Scan.perturb)
scRNA_Scan.interp_module <- Network_module(res_Scan.interp)
scRNA_SWEET_module <- Network_module(res_SWEET)
scRNA_SINUM_module <- Network_module(res_SINUM)

n_LIONESS <- length(scRNA_LIONESS_module)
scRNA_LIONESS_module_sim <- matrix(NA, n_LIONESS, n_LIONESS)
for (i in seq(n_LIONESS)) {
  for (j in seq(n_LIONESS)) {
    scRNA_LIONESS_module_sim[i, j] <- Sim.module(scRNA_LIONESS_module[[i]], scRNA_LIONESS_module[[j]])
  }
}

rownames(scRNA_LIONESS_module_sim) <- rownames(scRNA_LIONESS_net_sim)  
colnames(scRNA_LIONESS_module_sim) <- colnames(scRNA_LIONESS_net_sim) 

n_CSN <- length(scRNA_CSN_module)
scRNA_CSN_module_sim <- matrix(NA, n_CSN, n_CSN)
for (i in seq(n_CSN)) {
  for (j in seq(n_CSN)) {
    scRNA_CSN_module_sim[i, j] <- Sim.module(scRNA_CSN_module[[i]], scRNA_CSN_module[[j]])
  }
}

rownames(scRNA_CSN_module_sim) <- rownames(scRNA_CSN_net_sim)  
colnames(scRNA_CSN_module_sim) <- colnames(scRNA_CSN_net_sim) 

n_Scan.perturb <- length(scRNA_Scan.perturb_module)
scRNA_Scan.perturb_module_sim <- matrix(NA, n_Scan.perturb, n_Scan.perturb)
for (i in seq(n_Scan.perturb)) {
  for (j in seq(n_Scan.perturb)) {
    scRNA_Scan.perturb_module_sim[i, j] <- Sim.module(scRNA_Scan.perturb_module[[i]], scRNA_Scan.perturb_module[[j]])
  }
}

rownames(scRNA_Scan.perturb_module_sim) <- rownames(scRNA_Scan.perturb_net_sim)  
colnames(scRNA_Scan.perturb_module_sim) <- colnames(scRNA_Scan.perturb_net_sim) 

n_Scan.interp <- length(scRNA_Scan.interp_module)
scRNA_Scan.interp_module_sim <- matrix(NA, n_Scan.interp, n_Scan.interp)
for (i in seq(n_Scan.interp)) {
  for (j in seq(n_Scan.interp)) {
    scRNA_Scan.interp_module_sim[i, j] <- Sim.module(scRNA_Scan.interp_module[[i]], scRNA_Scan.interp_module[[j]])
  }
}

rownames(scRNA_Scan.interp_module_sim) <- rownames(scRNA_Scan.interp_net_sim)  
colnames(scRNA_Scan.interp_module_sim) <- colnames(scRNA_Scan.interp_net_sim) 

n_SWEET <- length(scRNA_SWEET_module)
scRNA_SWEET_module_sim <- matrix(NA, n_SWEET, n_SWEET)
for (i in seq(n_SWEET)) {
  for (j in seq(n_SWEET)) {
    scRNA_SWEET_module_sim[i, j] <- Sim.module(scRNA_SWEET_module[[i]], scRNA_SWEET_module[[j]])
  }
}

rownames(scRNA_SWEET_module_sim) <- rownames(scRNA_SWEET_net_sim)  
colnames(scRNA_SWEET_module_sim) <- colnames(scRNA_SWEET_net_sim) 

n_SINUM <- length(scRNA_SINUM_module)
scRNA_SINUM_module_sim <- matrix(NA, n_SINUM, n_SINUM)
for (i in seq(n_SINUM)) {
  for (j in seq(n_SINUM)) {
    scRNA_SINUM_module_sim[i, j] <- Sim.module(scRNA_SINUM_module[[i]], scRNA_SINUM_module[[j]])
  }
}

rownames(scRNA_SINUM_module_sim) <- rownames(scRNA_SINUM_net_sim)  
colnames(scRNA_SINUM_module_sim) <- colnames(scRNA_SINUM_net_sim) 

## Distribution of network, hub, and module similarity
install.packages("ggplot2")
library(ggplot2)

sim_list <- list(
  net   = scRNA_LIONESS_net_sim,
  hub   = scRNA_LIONESS_hub_sim,
  module = scRNA_LIONESS_module_sim
)

for (name in names(sim_list)) {
  vals <- sim_list[[name]][upper.tri(sim_list[[name]])]
  
  p <- ggplot(data.frame(similarity = vals), aes(x = similarity)) +
    geom_histogram(aes(y = after_stat(density)), bins = 10,
                   fill = "#0072B5", color = "white", alpha = 0.7) +
    geom_density(color = "#BC3C29", linewidth = 0.3) +
    labs(x = "Similarity", y = "Density") +
    theme_classic() +
    theme(text = element_text(size = 12), axis.title = element_text(size = 12),
          axis.text = element_text(size = 10), axis.line = element_line(linewidth = 0.5),
          panel.grid = element_blank())
  
  ggsave(paste0("similarity_distribution_", name, ".png"), p,
         width = 8, height = 6, dpi = 300, units = "cm")
}
```

# Clustering analysis
Clustering analysis was performed on the 19 K562 half-cells using an overall similarity matrix derived from the LIONESS method, which was calculated as the geometric mean of the network, hub, and module similarity matrices.

```{r, eval=TRUE, include=TRUE}
install.packages(c("umap", "ggplot2", "dplyr", "cluster"))
library(umap)
library(ggplot2)
library(dplyr)
library(cluster)

# Combined similarity matrix (geometric mean of three similarity matrices)
overall_sim <- exp((log(scRNA_LIONESS_net_sim) + log(scRNA_LIONESS_hub_sim) + log(scRNA_LIONESS_module_sim))/3)

# Convert similarity to distance matrix
dist_mat <- as.matrix(1 - overall_sim)
hc <- hclust(as.dist(dist_mat), method = "complete")

# Find optimal k (2-10) with all clusters >= 2
k_min <- 2
k_max <- min(10, nrow(dist_mat) - 1)
valid_ks <- c()
sil_vals <- c()
h_diffs <- c()

for (k in k_min:k_max) {
  cl <- cutree(hc, k = k)
  if (min(table(cl)) >= 2) {
    valid_ks <- c(valid_ks, k)
    sil_vals <- c(sil_vals, mean(silhouette(cl, as.dist(dist_mat))[, 3]))
    if (k < k_max) h_diffs <- c(h_diffs, diff(hc$height)[length(hc$height) - k])
    else h_diffs <- c(h_diffs, NA)
  }
}

if (length(valid_ks) == 0) {
  best_k <- 2
  warning("No k with all clusters >=2; forcing k=2")
} else {
  best_k_sil <- valid_ks[which.max(sil_vals)]
  best_k_elbow <- valid_ks[which.max(h_diffs[!is.na(h_diffs)])]
  best_k <- if (abs(best_k_sil - best_k_elbow) > 2) min(best_k_sil, best_k_elbow) else best_k_sil
}

# Cut tree at optimal k
clusters <- cutree(hc, k = best_k)

# Merge singletons (if any) to ensure all clusters >= 2
tab <- table(clusters)
if (min(tab) < 2) {
  for (s in as.numeric(names(tab[tab == 1]))) {
    idx <- which(clusters == s)
    d <- dist_mat[idx, ]
    d[idx] <- Inf
    clusters[idx] <- clusters[which.min(d)]
  }
  clusters <- as.numeric(factor(clusters))
}
cat("Final cluster sizes:\n"); print(table(clusters))

# UMAP dimensionality reduction
set.seed(123)
umap_out <- umap(d = dist_mat, input = "dist", n_neighbors = 6, min_dist = 0.3, n_components = 2)
df <- data.frame(
  UMAP1 = umap_out$layout[, 1],
  UMAP2 = umap_out$layout[, 2],
  Cluster = factor(clusters)
)

# Convex hulls and cluster centers
hulls <- df %>% group_by(Cluster) %>% filter(n() >= 3) %>% slice(chull(UMAP1, UMAP2))
centers <- df %>% group_by(Cluster) %>% summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2))

# Colors (up to 10 clusters)
cols <- c("#0072B5", "#E18727", "#20854E", "#BC3C29", "#7876B1",
          "#6A3D9A", "#FF7F00", "#FDBF6F", "#B15928", "#A6CEE3")[1:length(unique(clusters))]

p <- ggplot(df, aes(x = UMAP1, y = UMAP2, color = Cluster, fill = Cluster)) +
  geom_point(size = 1.0, alpha = 0.6) +
  geom_polygon(data = hulls, aes(fill = Cluster, color = Cluster),
               alpha = 0.15, linewidth = 0.3, show.legend = FALSE) +
  geom_text(data = centers, aes(label = Cluster), color = "black", size = 4, fontface = "bold") +
  scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
  theme_classic() +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(text = element_text(size = 10), axis.title = element_text(size = 10),
        axis.text = element_text(size = 8), axis.line = element_line(linewidth = 0.5),
        panel.grid = element_blank(), aspect.ratio = 1, legend.position = "right")

ggsave("umap_clustering.png", p, width = 10, height = 8, dpi = 300, units = "cm")
```

# Conclusions
In this tutorial, we apply S2N to show how to infer and validate sample-specific miRNA regulatory networks in the K562 dataset. Moreover, we also perform sample similarity analysis based on three key features derived from each sample-specific network, including network architecture, hub gene composition, and modular organization. Taken together, S2N provides a useful platform for advancing personalized biomedical research, from uncovering driver alterations in individual patients to stratifying cohorts based on sample-specific networks.

# Session information
```{r}
sessionInfo()
```
