---
title: "Tutorial for inferring sample-specific lncRNA regulatory networks with S2N in the TCGA-BRCA dataset" 
author: "\\
    
    Junpeng Zhang (zjp@dali.edu.cn)\\
    
    School of Engineering, Dali University"
date: '`r Sys.Date()`'
output:
    BiocStyle::html_document:
      toc: yes
    BiocStyle::pdf_document:
      toc: yes
vignette: >
    %\VignetteIndexEntry{Tutorial for inferring sample-specific lncRNA regulatory networks with S2N in the TCGA-BRCA dataset} 
    %\VignetteEngine{knitr::rmarkdown} 
    %\usepackage[utf8]{inputenc} 
    %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results="asis", message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
    warning = FALSE,
    message = FALSE)
```

# Introduction
In this tutorial, we apply 10 built-in methods (SSN, Paired-SSN, LIONESS, SCS, CSN, Scan.perturb, Scan.interp, SWEET, SINUM, and BONOBO) of S2N to identify tissue-specific lncRNA regulatory networks from the TCGA-BRCA dataset (https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-BRCA.star_fpkm-uq.tsv.gz). 

# Installation of S2N
S2N is freely available open-source under GNU GPL-3.0 from: https://github.com/zhangjunpeng411/S2N/, should be installed from GitHub prior to use.

```{r, eval=TRUE, include=TRUE}
install.packages("remotes")
library(remotes)
remotes::install_github("zhangjunpeng411/S2N")
library(S2N)
```

# Data preparation
The bulk gene expression data (TCGA-BRCA.star_fpkm-uq.tsv) are downloaded from UCSC Xena (http://xena.ucsc.edu/), and the processed data (log2(fpkm-uq+1)) and ID/Gene mapping information (gencode.v36.annotation.gtf.gene.probemap) are from https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-BRCA.star_fpkm-uq.tsv.gz, and https://gdc-hub.s3.us-east-1.amazonaws.com/download/gencode.v36.annotation.gtf.gene.probemap, respectively. Please paste all downloaded files into a single folder (set the folder as the directory of R environment).

```{r, eval=TRUE, include=TRUE}
gene_data <- read.delim("TCGA-BRCA.star_fpkm-uq.tsv", row.names = 1, stringsAsFactors = FALSE)
colnames(gene_data) <- gsub("\\.", "-", colnames(gene_data))
gene_probe_map <- read.delim("gencode.v36.annotation.gtf.gene.probemap", stringsAsFactors = FALSE)
```

We compute the average expression values of duplicate genes in the TCGA-BRCA dataset.

```{r, eval=TRUE, include=TRUE}
idx <- match(rownames(gene_data), gene_probe_map$id)
gene_names <- ifelse(is.na(idx), rownames(gene_data), gene_probe_map$gene[idx])
M <- as.matrix(gene_data)
group <- factor(gene_names, levels = unique(gene_names))
agg <- rowsum(M, group, reorder = FALSE)
sizes <- table(group)[rownames(agg)]
gene_data <- agg / as.vector(sizes)
```

The TCGA-BRCA dataset is further divided into two datssets (expression data of lncRNAs and mRNAs) according to the annotation information (hgnc_complete_set.txt) of HGNC (https://www.genenames.org/download/statistics-and-files/).

```{r, eval=TRUE, include=TRUE}
hgnc_complete_set <- read.delim("hgnc_complete_set.txt", stringsAsFactors=FALSE)
lncRNA_genes <- hgnc_complete_set$symbol[hgnc_complete_set$locus_type == "RNA, long non-coding"]
mRNA_genes <- hgnc_complete_set$symbol[hgnc_complete_set$locus_group == "protein-coding gene"]
lncRNA_data <- gene_data[rownames(gene_data) %in% lncRNA_genes, , drop = FALSE]
mRNA_data   <- gene_data[rownames(gene_data) %in% mRNA_genes, , drop = FALSE]
```

We only select matched tumor¨Cnormal pairs per patient.

```{r, eval=TRUE, include=TRUE}
barcodes <- colnames(lncRNA_data)
patient <- substr(barcodes, 1, 12)
code    <- as.integer(substr(barcodes, 14, 15))

is_tumor  <- code >= 1 & code <= 9
is_normal <- code >= 10 & code <= 19

paired_patients <- intersect(unique(patient[is_tumor]), unique(patient[is_normal]))
tumor_selected <- sapply(paired_patients, function(p) barcodes[patient == p & is_tumor][1])
normal_selected <- sapply(paired_patients, function(p) barcodes[patient == p & is_normal][1])
tumor_selected <- tumor_selected[order(paired_patients)]
normal_selected <- normal_selected[order(paired_patients)]
selected_cols <- c(tumor_selected, normal_selected)

lncRNA_data_map <- lncRNA_data[, selected_cols, drop = FALSE]
mRNA_data_map   <- mRNA_data[, selected_cols, drop = FALSE]
```

The TCGA-BRCA dataset is further preprocessed by removing duplicate and invariant genes, retaining only highly expressed and variable lncRNAs and mRNAs.

```{r, eval=TRUE, include=TRUE}
## Remove genes with zero expression values in all cells
lncRNA_RNA_norm_mean <- unlist(lapply(seq(dim(lncRNA_data_map)[1]), function(i) mean(lncRNA_data_map[i, ])))
lncRNA_RNA_norm_zero <- lncRNA_data_map[which(lncRNA_RNA_norm_mean > 0), ]
mRNA_RNA_norm_mean <- unlist(lapply(seq(dim(mRNA_data_map)[1]), function(i) mean(mRNA_data_map[i, ])))
mRNA_RNA_norm_zero <- mRNA_data_map[which(mRNA_RNA_norm_mean > 0), ]

## Reserve genes with higher mean expression values (top 10%) in all cells
lncRNA_mean <- rowMeans(lncRNA_RNA_norm_zero)
cutoff_lncRNA <- quantile(lncRNA_mean, probs = 0.9)
lncR_bulkRNA <- lncRNA_RNA_norm_zero[lncRNA_mean >= cutoff_lncRNA, ]
mRNA_mean <- rowMeans(mRNA_RNA_norm_zero)
cutoff_mRNA <- quantile(mRNA_mean, probs = 0.9)
mR_bulkRNA <- mRNA_RNA_norm_zero[mRNA_mean >= cutoff_mRNA, ]
lncR_bulkRNA_tumor <- lncR_bulkRNA[, 1:(ncol(lncR_bulkRNA)/2)]
lncR_bulkRNA_normal <- lncR_bulkRNA[, (ncol(lncR_bulkRNA)/2+1):ncol(lncR_bulkRNA)]
mR_bulkRNA_tumor <- mR_bulkRNA[, 1:(ncol(mR_bulkRNA)/2)]
mR_bulkRNA_normal <- mR_bulkRNA[, (ncol(mR_bulkRNA)/2+1):ncol(mR_bulkRNA)]
```

# Sample-specific lncRNA regulatory network construction
We use default parameters of 10 built-in methods (SSN, Paired-SSN, LIONESS, SCS, CSN, Scan.perturb, Scan.interp, SWEET, SINUM, and BONOBO) to infer tissue-specific lncRNA regulatory networks. In this context, each tissue represents a sample.

```{r, eval=TRUE, include=TRUE}
install.packages("peakRAM")
library(peakRAM)

dir.create("TCGA_BRCA_results", showWarnings = FALSE, recursive = TRUE)

exprs <- list(
  SSN = quote({
    res_SSN <- SSN(lncR_bulkRNA_normal, mR_bulkRNA_normal, lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_SSN, "TCGA_BRCA_results/bulkRNA_SSN_res.rds")
  }),
  PairedSSN = quote({
    res_PairedSSN <- PairedSSN(lncR_bulkRNA_normal, mR_bulkRNA_normal, lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_PairedSSN, "TCGA_BRCA_results/bulkRNA_PairedSSN_res.rds")
  }),
  LIONESS = quote({
    res_LIONESS <- LIONESS(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_LIONESS, "TCGA_BRCA_results/bulkRNA_LIONESS_res.rds")
  }),
  SCS = quote({
    res_SCS <- SCS(rownames(lncR_bulkRNA_tumor), lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_SCS, "TCGA_BRCA_results/bulkRNA_SCS_res.rds")
  }),
  CSN = quote({
    res_CSN <- CSN(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_CSN, "TCGA_BRCA_results/bulkRNA_CSN_res.rds")
  }),
  Scan.perturb = quote({
    res_Scan.perturb <- Scan.perturb(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, method = 'pearson', num.cores = 16)
    saveRDS(res_Scan.perturb, "TCGA_BRCA_results/bulkRNA_Scan.perturb_res.rds")
  }),
  Scan.interp = quote({
    res_Scan.interp <- Scan.interp(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, method = 'pearson', num.cores = 16)
    saveRDS(res_Scan.interp, "TCGA_BRCA_results/bulkRNA_Scan.interp_res.rds")
  }),
  SWEET = quote({
    res_SWEET <- SWEET(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_SWEET, "TCGA_BRCA_results/bulkRNA_SWEET_res.rds")
  }),
  SINUM = quote({
    res_SINUM <- SINUM(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_SINUM, "TCGA_BRCA_results/bulkRNA_SINUM_res.rds")
  }),
  BONOBO = quote({
    res_BONOBO <- BONOBO(lncR_bulkRNA_tumor, mR_bulkRNA_tumor, num.cores = 16)
    saveRDS(res_BONOBO, "TCGA_BRCA_results/bulkRNA_BONOBO_res.rds")
  })
)

peak_results <- do.call(peakRAM, exprs)
saveRDS(peak_results, "TCGA_BRCA_results/peakRAM_full_metrics_bulkRNA.rds")
```

# Sample-specific lncRNA regulatory network validation
To validate the predicted sample-specific lncRNA regulatory networks, we use established ground truth from miRTarBase v10.0 and TarBase v9.0.

```{r, eval=TRUE, include=TRUE}
data(lncRTarget_Groundtruth)
bulkRNA_SSN_validation <- Network_validation(res_SSN, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_PairedSSN_validation <- Network_validation(res_PairedSSN, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_LIONESS_validation <- Network_validation(res_LIONESS, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_CSN_validation <- Network_validation(res_CSN, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_Scan.perturb_validation <- Network_validation(res_Scan.perturb, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_Scan.interp_validation <- Network_validation(res_Scan.interp, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_SWEET_validation <- Network_validation(res_SWEET, lncRTarget_Groundtruth, calculate_metrics = TRUE)
bulkRNA_SINUM_validation <- Network_validation(res_SINUM, lncRTarget_Groundtruth, calculate_metrics = TRUE)

# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract performance values from each validation list
df_list <- list(
  SSN          = bulkRNA_SSN_validation[[2]][,4]*100,
  PairedSSN    = bulkRNA_PairedSSN_validation[[2]][,4]*100,
  LIONESS      = bulkRNA_LIONESS_validation[[2]][,4]*100,
  CSN          = bulkRNA_CSN_validation[[2]][,4]*100,
  Scan.perturb = bulkRNA_Scan.perturb_validation[[2]][,4]*100,
  Scan.interp  = bulkRNA_Scan.interp_validation[[2]][,4]*100,
  SWEET        = bulkRNA_SWEET_validation[[2]][,4]*100,
  SINUM        = bulkRNA_SINUM_validation[[2]][,4]*100
)

# Convert to long-format data frame (Method, Value)
df_long <- bind_rows(lapply(names(df_list), function(method) {
  data.frame(Method = method, Value = df_list[[method]])
}))

# Set Method as factor with the desired order
df_long$Method <- factor(df_long$Method, levels = names(df_list))

# Nature-inspired color palette 
science_colors <- c("#0072B5", "#E18727", "#20854E", "#BC3C29", 
                    "#7876B1", "#6A3D9A", "#FF7F00", "#FDBF6F")

# Generate boxplot with jittered points
p <- ggplot(df_long, aes(x = Method, y = Value, fill = Method)) +
  geom_boxplot(width = 0.6, alpha = 0.7, outlier.size = 0.5, 
               color = "black", outlier.alpha = 0.3) +
  geom_jitter(width = 0.15, size = 0.5, alpha = 0.3, color = "black") +
  scale_fill_manual(values = science_colors[1:length(unique(df_long$Method))]) +
  labs(x = "", y = "Percentage of validated lncRNA targets (%)") +
  theme_classic() +
  theme(text = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        axis.line = element_line(linewidth = 0.5, color = "black"),
        panel.grid = element_blank(),
        legend.position = "none")

# Display and save
print(p)
ggsave("performance_boxplot.png", p, width = 10, height = 9, dpi = 300, units = "cm")
```

# Sample similarity analysis
We calculate pairwise sample similarity based on three key features derived from each sample-specific network: (i) network architecture, (ii) hub gene composition, and (iii) modular organization. Histograms with kernel density overlay were generated from the BRCA tissue similarity matrices derived from sample-specific networks, hubs, and modules produced by the SWEET method.

```{r, eval=TRUE, include=TRUE}
# network-level
bulkRNA_SSN_net_sim <- Sim.network(res_SSN,  res_SSN)
bulkRNA_PairedSSN_net_sim <- Sim.network(res_PairedSSN,  res_PairedSSN)
bulkRNA_LIONESS_net_sim <- Sim.network(res_LIONESS,  res_LIONESS)
bulkRNA_CSN_net_sim <- Sim.network(res_CSN, res_CSN)
bulkRNA_Scan.perturb_net_sim <- Sim.network(res_Scan.perturb, res_Scan.perturb)
bulkRNA_Scan.interp_net_sim <- Sim.network(res_Scan.interp, res_Scan.interp)
bulkRNA_SWEET_net_sim <- Sim.network(res_SWEET, res_SWEET)
bulkRNA_SINUM_net_sim <- Sim.network(res_SINUM, res_SINUM)

# hub-level
bulkRNA_SSN_hub <- Network_hub(res_SSN)
bulkRNA_PairedSSN_hub <- Network_hub(res_PairedSSN)
bulkRNA_LIONESS_hub <- Network_hub(res_LIONESS)
bulkRNA_CSN_hub <- Network_hub(res_CSN)
bulkRNA_Scan.perturb_hub <- Network_hub(res_Scan.perturb)
bulkRNA_Scan.interp_hub <- Network_hub(res_Scan.interp)
bulkRNA_SWEET_hub <- Network_hub(res_SWEET)
bulkRNA_SINUM_hub <- Network_hub(res_SINUM)

bulkRNA_SSN_hub_sim <- Sim.hub(bulkRNA_SSN_hub,  bulkRNA_SSN_hub)
bulkRNA_PairedSSN_hub_sim <- Sim.hub(bulkRNA_PairedSSN_hub,  bulkRNA_PairedSSN_hub)
bulkRNA_LIONESS_hub_sim <- Sim.hub(bulkRNA_LIONESS_hub,  bulkRNA_LIONESS_hub)
bulkRNA_CSN_hub_sim <- Sim.hub(bulkRNA_CSN_hub, bulkRNA_CSN_hub)
bulkRNA_Scan.perturb_hub_sim <- Sim.hub(bulkRNA_Scan.perturb_hub, bulkRNA_Scan.perturb_hub)
bulkRNA_Scan.interp_hub_sim <- Sim.hub(bulkRNA_Scan.interp_hub, bulkRNA_Scan.interp_hub)
bulkRNA_SWEET_hub_sim <- Sim.hub(bulkRNA_SWEET_hub, bulkRNA_SWEET_hub)
bulkRNA_SINUM_hub_sim <- Sim.hub(bulkRNA_SINUM_hub, bulkRNA_SINUM_hub)

# module-level
bulkRNA_SSN_module <- Network_module(res_SSN)
bulkRNA_PairedSSN_module <- Network_module(res_PairedSSN)
bulkRNA_LIONESS_module <- Network_module(res_LIONESS)
bulkRNA_CSN_module <- Network_module(res_CSN)
bulkRNA_Scan.perturb_module <- Network_module(res_Scan.perturb)
bulkRNA_Scan.interp_module <- Network_module(res_Scan.interp)
bulkRNA_SWEET_module <- Network_module(res_SWEET)
bulkRNA_SINUM_module <- Network_module(res_SINUM)

n_SSN <- length(bulkRNA_SSN_module)
bulkRNA_SSN_module_sim <- matrix(NA, n_SSN, n_SSN)
for (i in seq(n_SSN)) {
  for (j in seq(n_SSN)) {
    bulkRNA_SSN_module_sim[i, j] <- Sim.module(bulkRNA_SSN_module[[i]], bulkRNA_SSN_module[[j]])
  }
}

rownames(bulkRNA_SSN_module_sim) <- rownames(bulkRNA_SSN_net_sim)  
colnames(bulkRNA_SSN_module_sim) <- colnames(bulkRNA_SSN_net_sim) 

n_PairedSSN <- length(bulkRNA_PairedSSN_module)
bulkRNA_PairedSSN_module_sim <- matrix(NA, n_PairedSSN, n_PairedSSN)
for (i in seq(n_PairedSSN)) {
  for (j in seq(n_PairedSSN)) {
    bulkRNA_PairedSSN_module_sim[i, j] <- Sim.module(bulkRNA_PairedSSN_module[[i]], bulkRNA_PairedSSN_module[[j]])
  }
}

rownames(bulkRNA_PairedSSN_module_sim) <- rownames(bulkRNA_PairedSSN_net_sim)  
colnames(bulkRNA_PairedSSN_module_sim) <- colnames(bulkRNA_PairedSSN_net_sim) 

n_LIONESS <- length(bulkRNA_LIONESS_module)
bulkRNA_LIONESS_module_sim <- matrix(NA, n_LIONESS, n_LIONESS)
for (i in seq(n_LIONESS)) {
  for (j in seq(n_LIONESS)) {
    bulkRNA_LIONESS_module_sim[i, j] <- Sim.module(bulkRNA_LIONESS_module[[i]], bulkRNA_LIONESS_module[[j]])
  }
}

rownames(bulkRNA_LIONESS_module_sim) <- rownames(bulkRNA_LIONESS_net_sim)  
colnames(bulkRNA_LIONESS_module_sim) <- colnames(bulkRNA_LIONESS_net_sim) 

n_CSN <- length(bulkRNA_CSN_module)
bulkRNA_CSN_module_sim <- matrix(NA, n_CSN, n_CSN)
for (i in seq(n_CSN)) {
  for (j in seq(n_CSN)) {
    bulkRNA_CSN_module_sim[i, j] <- Sim.module(bulkRNA_CSN_module[[i]], bulkRNA_CSN_module[[j]])
  }
}

rownames(bulkRNA_CSN_module_sim) <- rownames(bulkRNA_CSN_net_sim)  
colnames(bulkRNA_CSN_module_sim) <- colnames(bulkRNA_CSN_net_sim) 

n_Scan.perturb <- length(bulkRNA_Scan.perturb_module)
bulkRNA_Scan.perturb_module_sim <- matrix(NA, n_Scan.perturb, n_Scan.perturb)
for (i in seq(n_Scan.perturb)) {
  for (j in seq(n_Scan.perturb)) {
    bulkRNA_Scan.perturb_module_sim[i, j] <- Sim.module(bulkRNA_Scan.perturb_module[[i]], bulkRNA_Scan.perturb_module[[j]])
  }
}

rownames(bulkRNA_Scan.perturb_module_sim) <- rownames(bulkRNA_Scan.perturb_net_sim)  
colnames(bulkRNA_Scan.perturb_module_sim) <- colnames(bulkRNA_Scan.perturb_net_sim) 

n_Scan.interp <- length(bulkRNA_Scan.interp_module)
bulkRNA_Scan.interp_module_sim <- matrix(NA, n_Scan.interp, n_Scan.interp)
for (i in seq(n_Scan.interp)) {
  for (j in seq(n_Scan.interp)) {
    bulkRNA_Scan.interp_module_sim[i, j] <- Sim.module(bulkRNA_Scan.interp_module[[i]], bulkRNA_Scan.interp_module[[j]])
  }
}

rownames(bulkRNA_Scan.interp_module_sim) <- rownames(bulkRNA_Scan.interp_net_sim)  
colnames(bulkRNA_Scan.interp_module_sim) <- colnames(bulkRNA_Scan.interp_net_sim) 

n_SWEET <- length(bulkRNA_SWEET_module)
bulkRNA_SWEET_module_sim <- matrix(NA, n_SWEET, n_SWEET)
for (i in seq(n_SWEET)) {
  for (j in seq(n_SWEET)) {
    bulkRNA_SWEET_module_sim[i, j] <- Sim.module(bulkRNA_SWEET_module[[i]], bulkRNA_SWEET_module[[j]])
  }
}

rownames(bulkRNA_SWEET_module_sim) <- rownames(bulkRNA_SWEET_net_sim)  
colnames(bulkRNA_SWEET_module_sim) <- colnames(bulkRNA_SWEET_net_sim) 

n_SINUM <- length(bulkRNA_SINUM_module)
bulkRNA_SINUM_module_sim <- matrix(NA, n_SINUM, n_SINUM)
for (i in seq(n_SINUM)) {
  for (j in seq(n_SINUM)) {
    bulkRNA_SINUM_module_sim[i, j] <- Sim.module(bulkRNA_SINUM_module[[i]], bulkRNA_SINUM_module[[j]])
  }
}

rownames(bulkRNA_SINUM_module_sim) <- rownames(bulkRNA_SINUM_net_sim)  
colnames(bulkRNA_SINUM_module_sim) <- colnames(bulkRNA_SINUM_net_sim) 

## Distribution of network, hub, and module similarity
install.packages("ggplot2")
library(ggplot2)

sim_list <- list(
  net   = bulkRNA_SWEET_net_sim,
  hub   = bulkRNA_SWEET_hub_sim,
  module = bulkRNA_SWEET_module_sim
)

for (name in names(sim_list)) {
  vals <- sim_list[[name]][upper.tri(sim_list[[name]])]
  
  p <- ggplot(data.frame(similarity = vals), aes(x = similarity)) +
    geom_histogram(aes(y = after_stat(density)), bins = 20,
                   fill = "#0072B5", color = "white", alpha = 0.7) +
    geom_density(color = "#BC3C29", linewidth = 0.3) +
    labs(x = "Similarity", y = "Density") +
    theme_classic() +
    theme(text = element_text(size = 12), axis.title = element_text(size = 12),
          axis.text = element_text(size = 10), axis.line = element_line(linewidth = 0.5),
          panel.grid = element_blank())
  
  ggsave(paste0("similarity_distribution_", name, ".png"), p,
         width = 8, height = 6, dpi = 300, units = "cm")
}
```

# Clustering analysis
Clustering analysis was performed on the 113 BRCA tissues using an overall similarity matrix derived from the SWEET method, which was calculated as the geometric mean of the network, hub, and module similarity matrices.

```{r, eval=TRUE, include=TRUE}
install.packages(c("umap", "ggplot2", "dplyr", "cluster"))
library(umap)
library(ggplot2)
library(dplyr)
library(cluster)

# Combined similarity matrix (geometric mean of three similarity matrices)
overall_sim <- exp((log(bulkRNA_SWEET_net_sim) + log(bulkRNA_SWEET_hub_sim) + log(bulkRNA_SWEET_module_sim))/3)

# Convert similarity to distance matrix
dist_mat <- as.matrix(1 - overall_sim)
hc <- hclust(as.dist(dist_mat), method = "complete")

# Find optimal k (2-10) with all clusters >= 2
k_min <- 2
k_max <- min(10, nrow(dist_mat) - 1)
valid_ks <- c()
sil_vals <- c()
h_diffs <- c()

for (k in k_min:k_max) {
  cl <- cutree(hc, k = k)
  if (min(table(cl)) >= 2) {
    valid_ks <- c(valid_ks, k)
    sil_vals <- c(sil_vals, mean(silhouette(cl, as.dist(dist_mat))[, 3]))
    if (k < k_max) h_diffs <- c(h_diffs, diff(hc$height)[length(hc$height) - k])
    else h_diffs <- c(h_diffs, NA)
  }
}

if (length(valid_ks) == 0) {
  best_k <- 2
  warning("No k with all clusters >=2; forcing k=2")
} else {
  best_k_sil <- valid_ks[which.max(sil_vals)]
  best_k_elbow <- valid_ks[which.max(h_diffs[!is.na(h_diffs)])]
  best_k <- if (abs(best_k_sil - best_k_elbow) > 2) min(best_k_sil, best_k_elbow) else best_k_sil
}

# Cut tree at optimal k
clusters <- cutree(hc, k = best_k)

# Merge singletons (if any) to ensure all clusters >= 2
tab <- table(clusters)
if (min(tab) < 2) {
  for (s in as.numeric(names(tab[tab == 1]))) {
    idx <- which(clusters == s)
    d <- dist_mat[idx, ]
    d[idx] <- Inf
    clusters[idx] <- clusters[which.min(d)]
  }
  clusters <- as.numeric(factor(clusters))
}
cat("Final cluster sizes:\n"); print(table(clusters))

# UMAP dimensionality reduction
set.seed(123)
umap_out <- umap(d = dist_mat, input = "dist", n_neighbors = 30, min_dist = 0.3, n_components = 2)
df <- data.frame(
  UMAP1 = umap_out$layout[, 1],
  UMAP2 = umap_out$layout[, 2],
  Cluster = factor(clusters)
)

# Convex hulls and cluster centers
hulls <- df %>% group_by(Cluster) %>% filter(n() >= 3) %>% slice(chull(UMAP1, UMAP2))
centers <- df %>% group_by(Cluster) %>% summarise(UMAP1 = median(UMAP1), UMAP2 = median(UMAP2))

# Colors (up to 10 clusters)
cols <- c("#0072B5", "#E18727", "#20854E", "#BC3C29", "#7876B1",
          "#6A3D9A", "#FF7F00", "#FDBF6F", "#B15928", "#A6CEE3")[1:length(unique(clusters))]

p <- ggplot(df, aes(x = UMAP1, y = UMAP2, color = Cluster, fill = Cluster)) +
  geom_point(size = 1.0, alpha = 0.6) +
  geom_polygon(data = hulls, aes(fill = Cluster, color = Cluster),
               alpha = 0.15, linewidth = 0.3, show.legend = FALSE) +
  geom_text(data = centers, aes(label = Cluster), color = "black", size = 4, fontface = "bold") +
  scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
  theme_classic() +
  labs(x = "UMAP 1", y = "UMAP 2") +
  theme(text = element_text(size = 10), axis.title = element_text(size = 10),
        axis.text = element_text(size = 8), axis.line = element_line(linewidth = 0.5),
        panel.grid = element_blank(), aspect.ratio = 1, legend.position = "right")

ggsave("umap_clustering.png", p, width = 10, height = 8, dpi = 300, units = "cm")
```

# Conclusions
In this tutorial, we apply S2N to show how to infer and validate sample-specific lncRNA regulatory networks in the TCGA-BRCA dataset. Moreover, we also perform sample similarity analysis based on three key features derived from each sample-specific network, including network architecture, hub gene composition, and modular organization. Taken together, S2N provides a useful platform for advancing personalized biomedical research, from uncovering driver alterations in individual patients to stratifying cohorts based on sample-specific networks.

# Session information
```{r}
sessionInfo()
```
